---
title: "Overview"
output: 
  bookdown::pdf_document2: 
    toc: false
    number_sections: true
    citation_package: natbib
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{rmarkdown::pdf_document}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \pgfplotsset{compat=1.7}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \usepackage{mathtools}
  - \usepackage{dsfont}
  - \usepackage{prodint}
  - \usepackage{makecell}
  - \usepackage{booktabs}
---

<!-- output:  -->
<!--   rmarkdown::html_vignette: -->
<!--     number_sections: true -->



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(icmstate)
library(msm)
set.seed(1)
```


The `icmstate` package can be used to non-parametrically estimate the transition intensities in interval-censored multi-state models. 

In this vignette, we first give a brief [introduction](#Introduction) to [interval-censored data](#interval-censoring) and [multi-state modelling](#multi-state-modelling). [panel data](#panel-data) is intermittently observed data, which is well described by interval-censored multi-state models. Afterwards, we focus on the possible approaches for [non-parametric estimation](#Estimation) present in this package. We demonstrate the use of this package using an [example](#Example) and compare the estimates of this package to [known results](#Known-results). 


# Introduction {#Introduction}

## Interval-censoring {#interval-censoring}

The study of partially observed outcomes is called survival analysis. Outcomes are oftentimes partially observed due to a phenomenon known as right-censoring, where an outcomes is only known to be larger than a certain value. This often occurs in medical problems, where the survival time (after surgery, treatment, ...) of a patient is only known for a set of patients. This mathematical complication can happen for a number of reasons. Either patients are not followed up for a long enough time (study period was too short) or the patient is still alive or the patient was never inspected after a certain time period had passed due to other complications. The analysis of right-censored data is very well studied.

Another commonly encountered problem is interval-censoring. We speak of interval-censoring if the outcome is only known to lie between two values. An example of a setting where interval-censored data is collected is in the case of longitudinal follow-up. A patient might be assessed for the presence of a disease at multiple (scheduled) visits over a period of time. If the disease is detected at a certain visit, this means the patient has fallen ill between the previous two visit times. The exact time of infection is usually not known and nigh impossible to find out. 
Although data from such studies is clearly better described using interval-censoring, the time of infection is often assumed to be right-censored.


## Multi-state modelling {#multi-state-modelling}

Sometimes the available data is not well described by the simple survival setting, warranting the use of a more complicated multi-state model. In a multi-state model subjects can transition between different states, with the transition rate between states being of interest for estimation. The (extended) illness-death model is a famous example of a multi-state model, see Figure \@ref(fig:IDM).

\begin{figure}[!ht]
\centering
\begin{tikzpicture}
        \node (alive) [ draw, thick, rectangle, minimum width=2cm, minimum height = 1.5cm] at (0,0cm) {1. Alive};
        \node (illness) [ draw, thick, rectangle, minimum width=2cm, minimum height = 1.5cm] at (1.5,2cm) {2. Illness};
        \node (death) [ draw, thick, rectangle, minimum width=2cm, minimum height = 1.5cm] at (3,0cm)  {3. Death};
        \node (deathafterillness) [ draw, thick, dashed, rectangle, minimum width=2cm, minimum height = 1.5cm, text width=2cm] at (4.5,2cm) {4. Death after Illness};
        \draw[->, thick] (alive) -- (illness);
        \draw[->, thick, dotted] (illness) -- (death);
        \draw[->, thick, dashed] (illness) -- (deathafterillness);
        \draw[->, thick] (alive) -- (death);
\end{tikzpicture}
\caption{Graphical representation of the (extended) illness-death model. Extended: solid and dashed lines. Standard: solid and dotted lines.}
\label{fig:IDM}
\end{figure}






## Panel data {#panel-data}

When the state of subjects is observed intermittently, we obtain so called *panel data*. Such data is best described by an interval-censored multi-state model, as the exact transition times between states are not known. Additionally, it can be unclear exactly which transitions have been made between two observation times, as multi-state models can be quite complex. 

It sometimes occurs that the entry time into a subset of the states is always observed at an exact time. For example, in the illness-death model (Figure \@ref(fig:IDM)) it is reasonable to assume that the time of entry into the death state will be observed exactly. We will therefore also consider panel data with a mix of interval- and right-censored data.


# Estimation {#Estimation}

The `icmstate` package can be used to non-parametrically estimate the transition intensities in interval-censored Markov multi-state models. Let $\mathcal{H}$ denote the possible states in the model, so that $\mathcal{H} = \{1, 2, 3\}$ or $\mathcal{H} = \{\text{alive, illness, death}\}$ for the illness-death model. In general, the transition intensities are given by:
\begin{align*}
\alpha_{gh}(t) &= \lim_{dt \downarrow 0} \frac{\mathbb{P}(X(t + dt) = h | X(t) = g, \mathcal{F}_{t-})}{dt} = \lim_{dt \downarrow 0} \frac{\mathbb{P}(X(t + dt) = h | X(t) = g)}{dt},
\end{align*} with $X(t)$ the state of the process at time $t$ and $g, h \in \mathcal{H}$. Let $A_{gh}(t) = \int_0^t \alpha_{gh}(s) ds$ denote the cumulative transition intensity for the transition $g \to h$.

The transition probabilities are usually of greater interest than the intensities. These are defined as:
\begin{align*}
P_{gh}(s, t) = \mathbb{P}(X(t) = h | X(s) = g).
\end{align*} The transition intensities are related to the transition probabilities via product integration by the Chapman-Kolmogorov Equations:
\begin{align} \label{eq:transprobprodint}
    \mathbf{P}(s,t) = \Prodi_{s < u \leq t} \left( \mathbf{I} + \textrm{d}\mathbf{A}(u) \right),
\end{align} with $\mathbf{P}(s,t)$ the matrix containing the transition probabilities, with rows representing transitions from a state and columns transitions to a state. Letting $H = |\mathcal{H}|$, we have that $\mathbf{I}$ is the $H \times H$ identity matrix and $\textrm{d}\mathbf{A}(u)$ the matrix containing the transition intensities.


The theory above alows us to relate key quantities to each other. When estimating these quantities, some assumptions usually have to be made. Let us visualise a sample of $5$ subjects following an illness-death model (see Figure \@ref(fig:IDplot)).
```{r echo = FALSE, fig.keep='none'}
library(ggplot2)
library(latex2exp)
tmat <- mstate::trans.illdeath()

eval_times <- function(n_obs, stop_time){
  cumsum( c( 0,  runif( n_obs-1, 0, 2*(stop_time-4)/(n_obs-1) ) ) )
}
set.seed(1)
sim_dat <- sim_id_weib(n = 5, n_obs = 6, stop_time = 15, eval_times = eval_times,
                       start_state = "stable", shape = c(0.5, 0.5, 2), scale = c(5, 10, 10/gamma(1.5)))

ID_plot <- invisible(visualise_msm(remove_redundant_observations(sim_dat, tmat = tmat))$plot + xlab("Subject") + ylab("Time") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14,face="bold")) +
  ggtitle(label = "Illness-Death Model Data"))


unique_times <- unique(sort(c(ID_plot$data$t1, ID_plot$data$t2)))
n_unique <- length(unique_times)
# Create the vector of LaTeX expressions
tau_labels <- sapply(1:n_unique, function(i) TeX(paste0("$\\tau_{", i, "}$")))

ID_plot3 <- ID_plot + 
  geom_segment(aes(x = 0, xend = ID_plot$data$ID, y = ID_plot$data$t1, yend = ID_plot$data$t1), linetype = "dashed", col = "red", lwd = 1.3) + 
  geom_segment(aes(x = 0, xend = ID_plot$data$ID, y = ID_plot$data$t2, yend = ID_plot$data$t2), linetype = "dashed", col = "red", lwd = 1.3) + 
  scale_y_continuous(breaks = unique(sort(c(ID_plot$data$t1, ID_plot$data$t2))), labels = tau_labels)
```

```{r IDplot, echo = FALSE, fig.cap = "Visualisation of 5 subjects following an illness-death model. Each line represents the trajectory of a single subject, with the numbers indicating the state the subject was observed in at a certain time.", fig.align='center', out.width='70%'}
ID_plot 
```

In non-parametric estimation of simple interval-censored data, @Turnbull1976 had shown that it suffices to consider only the unique observation times of the subjects, and that the NPMLE of the cumulative intensities can only make jumps at these unique observation times. 

Although this hasn't been directly shown to be true for general interval-censored multi-state models, a similar assumption is usually made. We determine all unique observation times, and name them $\tau_1, \ldots, \tau_{K}$ with $K$ the total number of unique observation times. A visual representation with $K = 13$ of this can be seen in Figure \@ref(fig:IDplot3).

```{r IDplot3, echo = FALSE, fig.cap = "Unique observation times ordered chronologically and named.", fig.align='center', out.width='70%'}
ID_plot3 
```


Under this assumption, we are then interested in determining the jumps of the cumulative intensities only at these points in time. We therefore define:
\begin{align*}
\alpha_{gh}^k = \alpha_{gh}(\tau_k),
\end{align*} and this will be the main interest of our estimation procedure. Having obtained an estimate for these quantities, the Aalen-Johanssen estimator (Equation \@ref(eq:transprobprodint) with estimated intensities)  can be used to recover transition probabilities.


## Methods


The transition intensities are estimated using an EM algorithm. Two approaches are implemented in this package, described in the table below.


| Method          | Likelihood    | Exactly observed states  | R function |
| ----------------|---------| ------------| -------------- |
| @GomonPutter2024  | Multinomial| Implemented | `npmsm(..., method = "multinomial")` |
| @Gu2023        | Poisson    | Not implemented | `npmsm(..., method = "poisson")` |


The methods differ mostly in which complete-data likelihood is maximized, for details see the referenced articles. From here on out, we will refer to the two methods as Multinomial EM and Poisson EM.
Although the two methods usually yield similar estimates, there are some pros and cons to each method. 

* The Poisson EM estimator has been shown to be consistent when subjects can start in all non-absorbing states. This is not likely to be true for most panel data.
* The Multinomial EM algorithm requires way less iterations to converge to an estimate than the Poisson estimator. Due to this, computation times are usually more than twice as fast.
* The Multinomial EM algorithm allows for the incorporation of exactly observed states. Although possible with the Poisson EM algorithm as well, only a smooth approximation has been developed and is not incorporated in this package.
* The estimates for the transition probabilities have been shown to be consistent for both methods in a simulation study (@GomonPutter2024).

For most users, it should suffice to choose the appropriate method as follows:

1. If subjects can be initially observed in all non-absorbing states and the cumulative intensities are of interest: Poisson EM.
2. In all other cases, the computational advantage of the Multinomial EM algorithm will make it the more desirable choice.


<!-- \begin{table}[!ht] -->
<!-- \centering -->
<!-- \begin{tabular}{|c|c|c|} -->
<!-- \hline -->
<!--  & Multinomial & Poisson \\ \hline -->
<!-- pros & \makecell[c]{Fast \\ } & \makecell[c]{Consistency \\ } \\ \hline -->
<!-- cons & \makecell[c]{d \\ f} & \makecell[c]{asd} \\ \hline -->
<!-- \end{tabular} -->
<!-- \caption{A table with multiple rows in one cell.} -->
<!-- \end{table} -->









## Example {#Example}


We demonstrate basic usage of this package through a simulated example.


## Simple interval-censoring

The simplest form of a multi-state model is a model with two states. The NPMLE for this problem was found by @Turnbull1976. We show how to fit a simple model and compare our result with the well-known Turnbull estimator.

This ``standard'' survival setting can be represented by using the following *transition* matrix:

```{r}
library(mstate)
tmat <- transMat(x = list( c(2), c() ))
```

This transition matrix tells us that there are 2 states, with a possible transition to state $2$ from state $1$ and no transitions out of state $2$ (absorbing).

We want to generate some interval-censored data. This is possible by using the `msm` package (@Jackson2011). 
```{r}
library(msm)
#Absorbing state 2
qmatrix <- rbind(
  c(-0.2, 0.2),
  c(0, 0)
)
#Number of subjects in simulated data
n <- 10
#Create data frame for simulation:
simdat <- data.frame(time = c(replicate(n, c(0, seq(2, 12, by=2) + runif(6, 0, 2)))),
                    subject = rep(1:n, each = 7))
dat <- simmulti.msm(data = simdat, qmatrix = qmatrix, start = 1)[, 1:3]
names(dat)[1] <- "id"
```


We can visualise the data using the `visualise_msm()` function:
```{r}
visualise_msm(dat)
```
The visualisation will only show the observation times which are relevant for transitions, getting rid of repeated observations in the same state and observations in an absorbing state.

We transform the data for use by the `icenReg` package, so we can determine the Turnbull estimator
```{r}
icdata <- NULL
for(i in unique(dat$id)){
  gdi <- subset(dat, id == i)
  L_idx <- Position(isTRUE, gdi$state <= 1, right = TRUE)
  L <- gdi$time[L_idx] #Exit time from state 1
  if(L_idx < nrow(gdi)){
    R <- gdi$time[L_idx + 1]
  } else{
    R <- Inf
  }
  icdata <- rbind(icdata, c(L, R))
}
icdata <- as.data.frame(icdata)
colnames(icdata) <- c("L", "R")
```

Now we can determine the NPMLE using both methods:
```{r message=FALSE, warning=FALSE}
library(icenReg)
SimpleMult <- npmsm(gd = dat, tmat = tmat, method = "multinomial")
Turnbull <- ic_np(cbind(L, R) ~ 0, data = icdata)
```


Extract statistics to compare the two:
```{r}
Turnbull_result <- cbind(t(Turnbull$T_bull_Intervals), matrix(Turnbull$p_hat))
colnames(Turnbull_result) <- c("L", "R", "phat")
Turnbull_result
```
Unfortunately, the Turnbull estimator usually extracts $p_{12}^k = F(\tau_k) - F(\tau_{k-1})$ instead of the intensities. We can however compare the resulting survival functions of the two approaches:
```{r}
surv_turnbull <- sapply(1:nrow(Turnbull_result), function(i) 1 - sum(Turnbull_result[1:i, 3]))
surv_npmsm <- probtrans(SimpleMult$A, predt = 0, direction = "forward")
```




## Time homogeneous example





## Known results {#Known-results}

















# Additional Possibilities {#Additional-possibilities}

Although the non-parametric estimation of transition intensities was the main goal of this package, it has some additional possibilities which might be interesting to explore.


## Frydman (1995) non-parametric estimator {#Frydman-NPMLE}

The non-parametric estimator described in @Frydman1995 is available through the `msm_frydman()` function. It allows for the non-parametric estimation of a mix of cumulative distribution functions and cumulative intensity functions for the illness-death model with known transition paths. Details can be found in the article. This method can only be used if death times are observed exactly, and for each person it is known whether they have transitioned through illness on the way to death or not. It is therefore a very restricted version of the general NPMLE estimator described above. 

The support intervals (intervals where the NPMLE assigns positive probabilities) can be determined by feeding output from `msm_frydman()` to the `support_frydman()` function.


## Hudgens (2001/2005) support determination {#Hudgens-support}

@Hudgens2005 demonstrated an earlier result for determining the support intervals of interval censored data through the use of graph theory. He further extended the theory to determine when a NPMLE actually exists in the presence of truncation. The resulting theory can be used to determine existence of an NPMLE in the case of interval censored and truncated data using the `supportHudgens()` function. 

Additionally, an attempt was made to use the combination of @Hudgens2001 and @Hudgens2005 to determine the support of the general interval-censored multi-state problem. This attempt can be reproduced using the `estimate_support_msm()` function, although the resulting support intervals do not align with the empirical result of the `npmsm()` function. 
