---
title: "Overview"
output: 
  pdf_document: default
  number_sections: true
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

<!-- output:  -->
<!--   rmarkdown::html_vignette: -->
<!--     number_sections: true -->



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(icmstate)
```


The `icmstate` package can be used to non-parametrically estimate the transition intensities in interval-censored multi-state models. 

In this vignette, we first give a brief [introduction](#Introduction) to [interval-censored data](#interval-censoring) and [multi-state modelling](#multi-state-modelling). The combination of these two results in [panel data](#panel-data). Afterwards, we focus on the possible approaches for [non-parametric estimation](#Estimation) present in this package. We demonstrate the use of this package using an [example](#Example).


# Introduction {#Introduction}

## Interval-censoring {#interval-censoring}

The study of partially observed outcomes is called survival analysis. Outcomes are oftentimes partially observed due to a phenomenon known as right-censoring, where an outcomes is only known to be larger than a certain value. This often occurs in medical problems, where the survival time (after surgery, treatment, ...) of a patient is only known for a set of patients. This mathematical complication can happen for a number of reasons. Either patients are not followed up for a long enough time (study period was too short) or the patient is still alive or the patient was never inspected after a certain time period had passed due to other complications. The analysis of right-censored data is very well studied.

Another commonly encountered problem is interval-censoring. We speak of interval-censoring if the outcome is only known to lie between two values. An example of a setting where interval-censored data is collected is in the case of longitudinal follow-up. A patient might be assessed for the presence of a disease at multiple (scheduled) visits over a period of time. If the disease is detected at a certain visit, this means the patient has fallen ill between the previous two visit times. The exact time of infection is usually not known and nigh impossible to find out. 
Although data from such studies is clearly better described using interval-censoring, the time of infection is often assumed to be right-censored.


## Multi-state modelling {#multi-state-modelling}

Sometimes the available data is not well described by the simple survival setting, warranting the use of a more complicated multi-state model. In a multi-state model subjects can transition between different states, with the transition rate between states being of interest for estimation. The illness-death model is a famous example of a multi-state model, 


\begin{tikzpicture}
        \node (alive) [ draw, thick, rectangle, minimum width=2cm, minimum height = 1.5cm] at (0,0cm) {1. Alive};
        \node (illness) [ draw, thick, rectangle, minimum width=2cm, minimum height = 1.5cm] at (1.5,2cm) {2. Illness};
        \node (death) [ draw, thick, rectangle, minimum width=2cm, minimum height = 1.5cm] at (3,0cm)  {3. Death};
        \node (deathafterillness) [ draw, thick, dashed, rectangle, minimum width=2cm, minimum height = 1.5cm, text width=2cm] at (4.5,2cm) {4. Death after Illness};
        \draw[->, thick] (alive) -- (illness);
        \draw[->, thick, dotted] (illness) -- (death);
        \draw[->, thick, dashed] (illness) -- (deathafterillness);
        \draw[->, thick] (alive) -- (death);
\end{tikzpicture}

* More complicated data, not well described by simple survival setting.
* Introduce illness-death model, tikz picture

## Panel data {#panel-data}

# Estimation {#Estimation}




| Method          | Likelihood    | Exactly observed states  | R function |
| ----------------|---------| ------------| -------------- |
| @GomonPutter2024  | Multinomial| Implemented | `npmsm(, method = "binomial")` |
| @Gu2023        | Poisson    | Not implemented | `npmsm(, method = "poisson")` |




Write here about implementation of Frydman result from 1995.
Also write here about implementation of Hudgens results.


## Example {#Example}







