---
title: "ScenarioX_compact"
author: "Daniel Gomon"
date: '2024-06-26'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Scenario information - Adjust here to get appropriate output.

```{r}
scenario <- 4 #Choose from 1:4
n <- c(100, 300, 500) #Remove one if no files
n_obs <- c(6) #At most c(3,6)
N <- 1000
methods <- c("binomial", "msm")

eval_times <- seq(0, 15, 0.1)
w_shapes <- c(0.5, 0.5, 2)
w_scales <- c(5, 10, 10/gamma(1.5))
```



```{r echo = FALSE}
#Create file names to load:
load_names <- expand.grid(scenario, n, n_obs, N, methods)
colnames(load_names) <- c("scenario", "n", "n_obs", "N", "method")
load_names[, 5] <- as.character(load_names[, 5])
names_load <- c()
var_names <- c()
for(i in 1:nrow(load_names)){
  x <- load_names[i,]
  names_load <- c(names_load, paste0("sc", x[1], "n", x[2], "obs", trimws(x[3]), "N", x[4], x[5], ".Rdata"))
  var_names <- c(var_names, paste0("sc", x[1], "n", x[2], "obs", trimws(x[3]), "N", x[4], x[5]))
}
```


# Loading the data


```{r echo = FALSE}
for(i in 1:length(names_load)){
  load(names_load[i])
}
```


```{r}
if(inherits(get(var_names[1])[[1]], "npmsm")){
  n_states <- nrow(get(var_names[1])[[1]]$tmat) 
} else if(inherits(get(var_names[1])[[1]], "msm")){
  n_states <- get(var_names[1])[[1]]$qmodel$nstates
}

if(scenario == 3){
  shapes <- c(0.5, 0.5, 2)
  scales <- c(5, 10, 10/gamma(1.5))  
} else{
  shapes <- c(1, 1, 1)
  scales <- c(10, 20, 10)
}
from <- 1 #Which state do we consider transitions from?
```






# Loading the necessary packages

```{r echo = FALSE}
library(icmstate)
library(mstate)
library(ggplot2)
library(ggpubr)
```




# Cumulative Intensity Functions

## Function to determine interpolation of cumulative hazard, taking into account the support sets.

```{r echo = FALSE}
interpol_support <- function(msfit, trans){
  numeric_support <- support_npmsm(msfit)
  M <- length(numeric_support)
  if(trans > M){
    stop("Please provide a transition number which is possible.")
  }
  transno <- trans
  Haz <- rbind(c(0, 0, trans), subset(msfit$Haz, trans == transno))
  supp <- numeric_support[[trans]]$support[, c(1,2), drop = FALSE]
  
  if(nrow(supp) == 0){
    supp <- rbind(supp, c(min(Haz$time), max(Haz$time)))
  }
  
  #Make interpolation grid
  x <- matrix(t(supp), nrow = 1, byrow = FALSE)
  y <-  Haz$Haz[findInterval(x, Haz$time)]
  x <- c(0, x)
  y <- c(0, y)
  
  return(approxfun(x = x, y = y, rule = 2))
}

interpol_support_msm <- function(msm, trans){
  haz_rates <- exp(msm$estimates)
  return(function(t) ifelse(is.na(haz_rates[trans]), 0, haz_rates[trans] * t))
}
```




## Functions for time-specific extraction of statistics (RMSE, Bias, Variance)



```{r echo = FALSE}
#First we create a data-set where we have all the information we use to calculate summary statistics.
create_summary_df <- function(dat, eval_times){
  #Number of transitions
  if(inherits(dat[[1]], "npmsm")){
    M <- nrow(dat[[1]]$tmat2)  
  } else{
    M <- length(dat[[1]]$estimates)
  }
  
  #Number of evluation times
  n_times <- length(eval_times)
  #Number of replicates
  n_reps <- length(dat)
  df <- matrix(NA, nrow = n_reps * n_times, ncol = M + 2)
  for(i in 1:length(dat)){
    if(inherits(dat[[i]]$A, "msfit")){
      df[((i-1)*n_times +1):(i*n_times), 1:M] <- sapply(1:M, function(x) interpol_support(dat[[i]]$A, x)(eval_times))  
    } else{
      df[((i-1)*n_times +1):(i*n_times), 1:M] <- sapply(1:M, function(x) sapply(eval_times, function(t) interpol_support_msm(dat[[i]], x)(t)))   
    }
    df[((i-1)*n_times +1):(i*n_times), M+1] <- eval_times
    df[((i-1)*n_times +1):(i*n_times), M+2] <- rep(i, n_times)
  }
  colnames(df) <- c(paste0("trans", 1:M), "time", "id")
  return(df)
}
```



Now we can calculate summary statistics (RMSE, Bias, Variance):

```{r echo = FALSE}
extract_summary_stat <- function(summary_df, oracle_df, eval_times){
  M <- ncol(summary_df) - 2
  out_df <- NULL
  for(i in 1:length(eval_times)){
    current_time <- eval_times[i]
    relevant_df <- summary_df[summary_df[, "time"] == current_time, 1:M, drop = FALSE]
    relevant_oracle <- oracle_df[oracle_df[, "time"] == current_time, 1:M, drop = FALSE]
    #dat_trans contains difference between oracle hazard and estimated hazard for each transition (columns) for each repeat (rows)
    dat_trans <- t(apply(relevant_df, 1, function(x) x - relevant_oracle))
    #we therefore apply summary statistics to the columns to obtain them per transition
    stats <- c(t(apply(dat_trans, 2, function(x) c(mean(x), var(x), sqrt(var(x) + mean(x)^2)))))
    out_df <- rbind(out_df, data.frame(stats = stats, 
                                       type = rep(c("Bias", "Var", "RMSE"), each = 3),
                                       time = rep(current_time, M*3),
                                       trans = rep(1:M, 3)))
  }
  return(out_df)
}

```

We create an oracle:
```{r}
if(scenario != 3){ #Homogeneous oracle
  oracle_df <- matrix(NA, nrow = length(eval_times), ncol = 5)
  oracle_df[, 1] <- 0.1*eval_times
  oracle_df[, 2] <- 0.05*eval_times
  oracle_df[, 3] <- 0.1*eval_times
  oracle_df[, 4] <- eval_times
  oracle_df[, 5] <- rep(0, length(eval_times))
  colnames(oracle_df) <- c(paste0("trans", 1:3), "time", "id")  
} else{ #Weibull oracle
  oracle_df <- matrix(NA, nrow = length(eval_times), ncol = 5)
  oracle_df[, 1] <- -pweibull(eval_times, shape = w_shapes[1], scale = w_scales[1], lower = FALSE, log = TRUE)
  oracle_df[, 2] <- -pweibull(eval_times, shape = w_shapes[2], scale = w_scales[2], lower = FALSE, log = TRUE)
  oracle_df[, 3] <- -pweibull(eval_times, shape = w_shapes[3], scale = w_scales[3], lower = FALSE, log = TRUE)
  oracle_df[, 4] <- eval_times
  oracle_df[, 5] <- rep(0, length(eval_times))
  colnames(oracle_df) <- c(paste0("trans", 1:3), "time", "id")  
}


```






# Probability Transition Functions


```{r echo = FALSE}
#First we create a data-set where we have all the information we use to calculate summary statistics.
create_summary_df_pt <- function(dat, eval_times, from = from){
  #Number of transitions
  if(inherits(dat[[1]], "npmsm")){
    M <- nrow(dat[[1]]$tmat2)
    n_states <- nrow(dat[[1]]$tmat)
  } else{
    M <- dat[[1]]$qmodel$npars
    n_states <- dat[[1]]$qmodel$nstates
  }
  
  #Number of evluation times
  n_times <- length(eval_times)
  #Number of replicates
  n_reps <- length(dat)
  df <- matrix(NA, nrow = n_reps * n_times, ncol = n_states + 2)
  for(i in 1:length(dat)){
    if(inherits(dat[[i]]$A, "msfit")){
      df[((i-1)*n_times +1):(i*n_times), 1:(n_states+1)] <- suppressWarnings({
      as.matrix(probtrans(interpol_msfit(dat[[i]]$A, times = eval_times), predt = min(eval_times), direction = "forward", variance = FALSE)[[from]])
      })
    } else{
      df[((i-1)*n_times +1):(i*n_times), 1:(n_states+1)] <- suppressWarnings({
        as.matrix(probtrans_msm(dat[[i]], times = eval_times)[[from]])
      })
    }
    df[((i-1)*n_times +1):(i*n_times), n_states+2] <- rep(i, n_times)
  }
  colnames(df) <- c("time", paste0("tostate", 1:n_states), "id")
  return(df)
}
```



Now we can calculate summary statistics (RMSE, Bias, Variance):

```{r echo = FALSE}
extract_summary_stat_pt <- function(summary_df, oracle_df, eval_times){
  M <- n_states <- ncol(summary_df) - 2
  out_df <- NULL
  for(i in 1:length(eval_times)){
    current_time <- eval_times[i]
    relevant_df <- summary_df[summary_df[, "time"] == current_time, 2:(n_states+1), drop = FALSE]
    relevant_oracle <- as.matrix(oracle_df[oracle_df[, "time"] == current_time, 2:(n_states+1), drop = FALSE])
    #dat_trans contains difference between oracle hazard and estimated hazard for each transition (columns) for each repeat (rows)
    dat_trans <- t(apply(relevant_df, 1, function(x) x - relevant_oracle))
    #we therefore apply summary statistics to the columns to obtain them per transition
    stats <- c(t(apply(dat_trans, 2, function(x) c(mean(x), var(x), sqrt(var(x) + mean(x)^2)))))
    out_df <- rbind(out_df, data.frame(stats = stats, 
                                       type = rep(c("Bias", "Var", "RMSE"), each = n_states),
                                       time = rep(current_time, M*3),
                                       trans = rep(1:n_states, 3)))
  }
  return(out_df)
}

```


#Extract Data

## Intensities 


```{r}
for(i in 1:nrow(load_names)){
  assign(paste0("summary_df", i), suppressWarnings(create_summary_df(get(var_names[i]), eval_times = eval_times)))
}
for(i in 1:nrow(load_names)){
  assign(paste0("stat_df", i), extract_summary_stat(summary_df = get(paste0("summary_df", i)), oracle_df = oracle_df, eval_times = eval_times))
}
```



## Transition Probabilities


```{r}
oracle_df_pt <- probtrans_weib(transMat = get(var_names[1])[[1]]$tmat, times = eval_times, shapes = shapes, scales = scales, type = "prodint")[[from]]

for(i in 1:nrow(load_names)){
  assign(paste0("summary_df_pt", i), suppressWarnings(create_summary_df_pt(get(var_names[i]), eval_times = eval_times)))
}
for(i in 1:nrow(load_names)){
  assign(paste0("stat_df_pt", i), extract_summary_stat_pt(summary_df = get(paste0("summary_df_pt", i)), oracle_df = oracle_df_pt, eval_times = eval_times))
}
```






# Plotting 


## Cumulative Intensities

Compare the statistics as n increases within methods.

```{r fig.keep='all'}
trans_names <- c("1.Alive -> 2.Illness", "1.Alive -> 3.Death", "2.Illness -> 4.Death (aI)")
names(trans_names) <- c(1, 2, 3)
method_names <- c("Multinomial EM", "Homogeneous")
names(method_names) <- c("binomial", "msm")


for(i in 1:length(methods)){
  stat_multiple <- NULL
  for(j in 1:length(n)){
    stat_multiple <- rbind(stat_multiple, get(paste0("stat_df", (i-1)*length(n)+j)))
  }
  stat_multiple <- cbind(stat_multiple, c(rep(n, each = length(eval_times) * 9)))
  colnames(stat_multiple)[5] <- "n"
  plot_all_stat <- ggplot(data = stat_multiple, aes(x = time, y = stats, group = interaction(type, n), col = as.factor(n), linetype = as.factor(type))) + geom_line(lwd = 1.2) + facet_grid(trans~., scales = "free", labeller = as_labeller(trans_names)) + xlab("Time (Years)") + ylab("Value") + labs(title = paste0("Cumulative Intensities"), subtitle = paste0("Scenario ", scenario, ", ", method_names[i]), color = "Sample size", linetype = "Measure") + scale_color_brewer(palette="Dark2")
  print(plot_all_stat)
  assign(paste0("plot_all_statn", i), plot_all_stat)
}

all_stat_plotn <- ggarrange(plot_all_statn1, plot_all_statn2,
          labels = c("A", "B"),
          ncol = 3, nrow = 1, common.legend = TRUE, legend = "bottom")


ggsave(file = paste0("Scenario", scenario, "gg_compint.eps"), plot = all_stat_plotn, width = 5.5, height = 3, units = "in", scale = 1.7)
```


Compare between methods for fixed n.


```{r}
trans_names <- c("1.Alive -> 2.Illness", "1.Alive -> 3.Death", "2.Illness -> 4.Death (aI)")
names(trans_names) <- c(1, 2, 3)
method_names <- c("Multinomial EM", "Homogeneous")
names(method_names) <- c("binomial", "msm")


for(i in 1:length(n)){
  stat_multiple <- NULL
  for(j in 1:length(methods)){
    stat_multiple <- rbind(stat_multiple, get(paste0("stat_df", i + length(n)*(j-1))))
  }
  stat_multiple <- cbind(stat_multiple, c(rep(methods, each = length(eval_times) * 9)))
  colnames(stat_multiple)[5] <- "Method"
  plot_all_stat <- ggplot(data = stat_multiple, aes(x = time, y = stats, group = interaction(type, as.factor(Method)), col = as.factor(Method), linetype = as.factor(type))) + geom_line(lwd = 1.2) + facet_grid(trans~., scales = "free", labeller = as_labeller(trans_names)) + xlab("Time (Years)") + ylab("Value") + labs(title = "Cumulative Intensities", subtitle = paste0("Scenario ", scenario,  ", n = ", n[i]), color = "Method", linetype = "Measure") + scale_color_brewer(palette="Dark2", labels = method_names)
  print(plot_all_stat)
  assign(paste0("plot_all_stat", i), plot_all_stat)
}

all_stat_plot <- ggarrange(plot_all_stat1, plot_all_stat2, plot_all_stat3, 
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1, common.legend = TRUE, legend = "bottom")


ggsave(file = paste0("Scenario", scenario, "gg_intensities.eps"), plot = all_stat_plot, width = 5.5, height = 3, units = "in", scale = 1.7)

```






## Transition Probabilities


Compare the statistics as n increases within methods.

```{r fig.keep='all'}

trans_names <- c("1.Alive -> 1.Alive", "1.Alive -> 2.Illness", "1.Alive -> 3.Death", "1.Alive -> 4.Death (aI)")
names(trans_names) <- c(1, 2, 3, 4)
method_names <- c("Multinomial EM", "Poisson EM", "Homogeneous")
names(method_names) <- c("binomial", "poisson", "msm")

for(i in 1:length(methods)){
  stat_multiple <- NULL
  for(j in 1:length(n)){
    stat_multiple <- rbind(stat_multiple, get(paste0("stat_df_pt", (i-1)*length(n)+j)))
  }
  stat_multiple <- cbind(stat_multiple, c(rep(n, each = length(eval_times) * n_states * length(n))))
  colnames(stat_multiple)[5] <- "n"
  plot_all_stat_ptn <- ggplot(data = stat_multiple, aes(x = time, y = stats, group = interaction(type, n), col = as.factor(n), linetype = as.factor(type))) + geom_line(lwd = 1.2) + facet_grid(trans~., scales = "free", labeller = as_labeller(trans_names)) + xlab("Time (Years)") + ylab("Value") + labs(title = "Transition Probabilities", subtitle = paste0("Scenario ", scenario,  ", ", method_names[i]), color = "Sample size", linetype = "Measure") + scale_color_brewer(palette="Dark2", labels = method_names) 
  print(plot_all_stat_ptn)
  assign(paste0("plot_all_stat_ptn", i), plot_all_stat_ptn)
}

all_stat_plot_ptn <- ggarrange(plot_all_stat_ptn1, plot_all_stat_ptn2,
          labels = c("A", "B"),
          ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")

ggsave(file = paste0("Scenario", scenario, "gg_compprobs.eps"), plot = all_stat_plot_ptn, width = 5.5, height = 3, units = "in", scale = 1.7)
```

Compare between methods for fixed n.


```{r}
trans_names <- c("1.Alive -> 1.Alive", "1.Alive -> 2.Illness", "1.Alive -> 3.Death", "1.Alive -> 4.Death (aI)")
names(trans_names) <- c(1, 2, 3, 4)
method_names <- c("Multinomial EM", "Poisson EM", "Homogeneous")
names(method_names) <- c("binomial", "poisson", "msm")



for(i in 1:length(n)){
  stat_multiple <- NULL
  for(j in 1:length(methods)){
    stat_multiple <- rbind(stat_multiple, get(paste0("stat_df_pt", i + length(n)*(j-1))))
  }
  stat_multiple <- cbind(stat_multiple, c(rep(methods, each = length(eval_times) * n_states * 3)))
  colnames(stat_multiple)[5] <- "Method"
  plot_all_stat_pt <- ggplot(data = stat_multiple, aes(x = time, y = stats, group = interaction(type, as.factor(Method)), col = as.factor(Method), linetype = as.factor(type))) + geom_line(lwd = 1.2) + facet_grid(trans~., scales = "free", labeller = as_labeller(trans_names)) + xlab("Time (Years)") + ylab("Value") + labs(title = "Transition Probabilities", subtitle = paste0("Scenario ", scenario,  ", n = ", n[i]), color = "Method", linetype = "Measure") + scale_color_brewer(palette="Dark2", labels = method_names) 
  print(plot_all_stat_pt)
  assign(paste0("plot_all_stat_pt", i), plot_all_stat_pt)
}

all_stat_plot_pt <- ggarrange(plot_all_stat_pt1, plot_all_stat_pt2, plot_all_stat_pt3, 
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1, common.legend = TRUE, legend = "bottom")


ggsave(file = paste0("Scenario", scenario, "gg_probs.eps"), plot = all_stat_plot_pt, width = 5.5, height = 3, units = "in", scale = 1.7)

```







## Combination plots

for n = 500, plot cumulative intensities and transition probabilities

```{r}
trans_names <- c("1.Alive -> 1.Alive", "1.Alive -> 2.Illness", "1.Alive -> 3.Death", "1.Alive -> 4.Death (aI)")
names(trans_names) <- c(1, 2, 3, 4)
method_names <- c("Multinomial EM", "Poisson EM", "Homogeneous")
names(method_names) <- c("binomial", "poisson", "msm")


#only for n = 500
k <- 3


stat_multiple <- NULL
for(j in 1:length(methods)){
  stat_multiple <- rbind(stat_multiple, get(paste0("stat_df_pt", k + length(n)*(j-1))))
}
stat_multiple <- cbind(stat_multiple, c(rep(methods, each = length(eval_times) * n_states * 3)))
colnames(stat_multiple)[5] <- "Method"
comb_plot <- ggplot(data = stat_multiple, aes(x = time, y = stats, group = interaction(type, as.factor(Method)), col = as.factor(Method), linetype = as.factor(type))) + geom_line(lwd = 1.2) + facet_grid(trans~., scales = "free", labeller = as_labeller(trans_names)) + xlab("Time (Years)") + ylab("Value") + labs(title = "Transition Probabilities", subtitle = paste0("Scenario ", scenario,  ", n = ", n[i]), color = "Method", linetype = "Measure") + scale_color_brewer(palette="Dark2", labels = method_names) 



trans_names <- c("1.Alive -> 2.Illness", "1.Alive -> 3.Death", "2.Illness -> 4.Death (aI)")
names(trans_names) <- c(1, 2, 3)
method_names <- c("Multinomial EM", "Homogeneous")
names(method_names) <- c("binomial", "msm")



stat_multiple <- NULL
for(j in 1:length(methods)){
  stat_multiple <- rbind(stat_multiple, get(paste0("stat_df", k + length(n)*(j-1))))
}
stat_multiple <- cbind(stat_multiple, c(rep(methods, each = length(eval_times) * 9)))
colnames(stat_multiple)[5] <- "Method"
comb_plot2 <- ggplot(data = stat_multiple, aes(x = time, y = stats, group = interaction(type, as.factor(Method)), col = as.factor(Method), linetype = as.factor(type))) + geom_line(lwd = 1.2) + facet_grid(trans~., scales = "free", labeller = as_labeller(trans_names)) + xlab("Time (Years)") + ylab("Value") + labs(title = "Cumulative Intensities", subtitle = paste0("Scenario ", scenario,  ", n = ", n[i]), color = "Method", linetype = "Measure") + scale_color_brewer(palette="Dark2", labels = method_names)






all_comb_plot <- ggarrange(comb_plot, comb_plot2, 
          labels = c("A", "B"),
          ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")

ggsave(file = paste0("Scenario", scenario, "gg_combplot.eps"), plot = all_comb_plot, width = 5.5, height = 3, units = "in", scale = 1.7)

```



